---
title: "Daisy Metabolomics analysis"
author: "Pei Qin (Sabrina) Ng"
date: "11/06/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Samples are H-RPE cells treated with TGFb2 or TNFa (cell + media metabolites)


Datasets were reformatted for analysis using Omu package
```{r}
library(readxl)
library(dplyr)
library(tibble)
library(ggplot2)
library(viridis)
library(tibble)

data <- read_xlsx("mx_608189_Daisy Shu_human cells_ media_primary_metabolism_GC-TOF_MS_10-2021_submit.xlsx", sheet = "sub")
metadata <- read_xlsx("mx_608189_Daisy Shu_human cells_ media_primary_metabolism_GC-TOF_MS_10-2021_submit.xlsx", sheet = "metadata")

data <- data %>% 
        rename(Metabolite = "BinBase name")

#export as csv format
write.csv(data, "metabolics_raw_data.csv")

#transpose the original metadata format to suit omu package requirements

metadata <- t(metadata) %>% as.data.frame() %>% 
    tibble::rownames_to_column("mx_Lane") %>%
    select(V1,V3,V4,V2,V5) %>% 
    tidyr::unite(Background, c(V3,V4)) %>% 
    rename(Sample = V1, Treatment = V2, Grouped =V5)

write.csv(metadata, "metabolics_metadata.csv")
```

Reformatted data were used as input (CSV data) into Omu.

```{r}
library(omu)
library(dplyr)
library(tibble)

metabolomics_count_dataframe <- read_metabo(filepath = "metabolics_raw_data.csv")
metabolomics_metadata <- read.csv("metabolics_metadata.csv") #included the KEGG ID for cadaverine and other missing primary metabolites manually (those remained NA means no record)

metabolomics_metadata <- mutate_if(metabolomics_metadata, 
                                   is.character, as.factor) %>%
                        subset(select = -c(X)) %>%as.data.frame() %>%
                        filter(Sample != "X10_010")

metabolomics_count_dataframe_all  <- metabolomics_count_dataframe %>% 
                                     subset(select = -c(X,ret.index,
                                                   quant.mz,
                                                   BB.id,
                                                   mass.spec,
                                                   PubChem,
                                                   InChI.Key,
                                                   cells,media)) %>% 
                                                    as.data.frame()

#data reformatting for the purpose of PCA plot. Note PCAplot function for omu does not work too well with this dataset.
metabolomics_count_dataframe  <- metabolomics_count_dataframe %>% 
                                subset(select = -c(X,ret.index,
                                                   quant.mz,
                                                   BB.id,
                                                   mass.spec,
                                                   PubChem,
                                                   InChI.Key,
                                                   cells,media)) %>% as.tibble() %>% 
                                    select(Metabolite,X1_001:X48_048) %>%
                                    tidyr::pivot_longer(X1_001:X48_048, 
                                     names_to = "Sample", values_to = "peaks" ) %>% 
                                     tidyr::pivot_wider(names_from = Metabolite, values_from = peaks)





rownames(metabolomics_metadata) <- NULL

comb_data <- left_join(metabolomics_count_dataframe,
                       metabolomics_metadata, 
                       by = "Sample")


#Samples which are human cells 
cells_list <- comb_data %>% 
              filter(Background == "human_cells") %>% 
              distinct(Sample)

cells_counts_Df <- metabolomics_count_dataframe  %>% 
                    filter(Sample %in% cells_list$Sample) %>%
                 column_to_rownames("Sample")

#for annotation
cells_comb_data <- left_join(metabolomics_count_dataframe,
                       metabolomics_metadata, 
                       by = "Sample") %>% 
    relocate (Sample, Treatment) %>% 
             filter(Sample %in% cells_list$Sample)

#Samples which are media 
media_list <- comb_data %>% 
              filter(Background == "human_media") %>% 
              distinct(Sample)

media_counts_Df <- metabolomics_count_dataframe  %>% 
                    filter(Sample %in% media_list$Sample) %>%
                 column_to_rownames("Sample")

#for annotation
media_comb_data <- left_join(metabolomics_count_dataframe,
                       metabolomics_metadata, 
                       by = "Sample") %>% 
                   relocate (Sample, Treatment) %>% 
                  filter(Sample %in% media_list$Sample)

#only use up till this stage as this package is pretty useless
```


Split the data by:

1) Cell lines

2) media

PCA plots for each of these group in cells
```{r}
pca_res <- prcomp(sqrt(cells_counts_Df), center = TRUE, scale. = TRUE)

df <-as.data.frame(pca_res$x) %>% rownames_to_column("Sample") %>% mutate(Treatments = cells_comb_data$Treatment) %>% select(Sample,Treatments, PC1:PC23)


ggplot(df, aes(PC1, PC2, colour = factor(Treatments))) + 
  geom_point(size = 3) +
  theme_bw() + labs(colour = "Groups")
  
#PCA_plot(count_data = cells_counts_log_Df, metadata = metabolomics_metadata, variable = "Treatment", color = "Treatment", response_variable = "Metabolites")+ theme_bw() + theme(panel.grid = element_blank())
```

PCA plots for each of these group in media
```{r}
pca_res_2 <- prcomp(sqrt(media_counts_Df), center = TRUE, scale. = TRUE)
df_2 <- as.data.frame(pca_res_2$x) %>% rownames_to_column("Sample") %>% mutate(Treatments = media_comb_data$Treatment) %>% select(Sample,Treatments, PC1:PC23)


ggplot(df_2, aes(PC1, PC2, colour = factor(Treatments))) + 
  geom_point(size = 3) +
  theme_bw() + labs(colour = "Groups")
```

Assigns hierarchy metadata to a metabolomics count matrix using identifier values. 
It can assign KEGG compound hierarchy, orthology hierarchy, or organism hierarchy data.

1) Cells_TNFa vs Cells_Control
```{r}
DF <- assign_hierarchy(metabolomics_count_dataframe_all, 
                       keep_unknowns = FALSE, 
                       identifier = "KEGG" ) 

#remove sample 10 from the data

DF <- DF %>% subset(select=-c(X10_010))

DF_stats_ca <-omu_summary(count_data = DF , 
            metadata = metabolomics_metadata, 
            numerator = "Cells_TNFa", 
            denominator = "Cells_Control", 
            response_variable = "Metabolite", 
            Factor = "Treatment", 
            p_adjust = "BH", 
            test_type = "welch")


```

Generate stats of differential comparison of metabolites detected in Cells_TNFa vs Cells_Control

```{r}
DF_stats_counts_ca <- count_fold_changes(count_data = DF_stats_ca, 
                                      column = "Class", 
                                      sig_threshold = 0.05, 
                                      keep_unknowns = FALSE)

writexl::write_xlsx(DF_stats_counts_ca, "DifferentialMetabolites_Cells_TNFavsControl.xlsx")
```

Visualise results for Cells_TNFa vs Cells_control in barplot to show comparisons of metabolites

```{r}
library(ggplot2)
Class_Bar_Plot_1 <- plot_bar(fc_data = DF_stats_counts_ca, 
                           fill = c("dodgerblue2", "firebrick2"), 
                           outline_color = c("black", "black"), size = c(1,1)) + 
                  labs(x = "Class") + 
                  theme(panel.grid = element_blank()) + 
                ggtitle("Metabolites comparison between Cells_TNFa vs Cells_Control")
```


Repeat the analysis for the following comparisons

2)Cells_TGFb vs Cells_Control

```{r}

DF_stats_cb <-omu_summary(count_data = DF , 
            metadata = metabolomics_metadata, 
            numerator = "Cells_TGFb", 
            denominator = "Cells_Control", 
            response_variable = "Metabolite", 
            Factor = "Treatment", 
            p_adjust = "BH", 
            test_type = "welch")

DF_stats_counts_cb <- count_fold_changes(count_data = DF_stats_cb, 
                                      column = "Class", 
                                      sig_threshold = 0.05, 
                                      keep_unknowns = FALSE)

writexl::write_xlsx(DF_stats_counts_cb, "DifferentialMetabolites_Cells_TGFbvsControl.xlsx")

Class_Bar_Plot_2 <- plot_bar(fc_data = DF_stats_counts_cb, 
                           fill = c("dodgerblue2", "firebrick2"), 
                           outline_color = c("black", "black"), size = c(1,1)) + 
                  labs(x = "Class") + 
                  theme(panel.grid = element_blank()) + 
                ggtitle("Metabolites comparison between Cells_TGFb vs Cells_Control")
```


3) Media_TNFa vs Media_Control
```{r}
DF_stats_ma <-omu_summary(count_data = DF , 
            metadata = metabolomics_metadata, 
            numerator = "Media_TNFa", 
            denominator = "Media_Con", 
            response_variable = "Metabolite", 
            Factor = "Treatment", 
            p_adjust = "BH", 
            test_type = "welch")

DF_stats_counts_ma <- count_fold_changes(count_data = DF_stats_ma, 
                                      column = "Class", 
                                      sig_threshold = 0.05, 
                                      keep_unknowns = FALSE)

Class_Bar_Plot_3 <- plot_bar(fc_data = DF_stats_counts_ma, 
                           fill = c("dodgerblue2", "firebrick2"), 
                           outline_color = c("black", "black"), size = c(1,1)) + 
                  labs(x = "Class") + 
                  theme(panel.grid = element_blank()) + 
                ggtitle("Metabolites comparison between Media_TNFa vs Media_Control")
```

4) Media_TGFb vs Media_Control
```{r}
DF_stats_mb <-omu_summary(count_data = DF , 
            metadata = metabolomics_metadata, 
            numerator = "Media_TGFb", 
            denominator = "Media_Con", 
            response_variable = "Metabolite", 
            Factor = "Treatment", 
            p_adjust = "BH", 
            test_type = "welch")

DF_stats_counts_mb <- count_fold_changes(count_data = DF_stats_mb, 
                                      column = "Class", 
                                      sig_threshold = 0.05, 
                                      keep_unknowns = FALSE)

Class_Bar_Plot_4 <- plot_bar(fc_data = DF_stats_counts_mb, 
                           fill = c("dodgerblue2", "firebrick2"), 
                           outline_color = c("black", "black"), size = c(1,1)) + 
                  labs(x = "Class") + 
                  theme(panel.grid = element_blank()) + 
                ggtitle("Metabolites comparison between Media_TGFb vs Media_Control")
```
For both media comparisons, no significant changes of metabolites detected.


Volcano plots (For Cells TNFa and Control, NAs included)

```{r}
#generate the variables required for the volcano plot
list_ca <- unique(DF_stats_ca$Class) %>% as.vector()
pal <- viridis(option = "C", 10)

#top 10 labels
#with GG plot the amount of roles in the label you want have to still match the dataframe which you are plotting
#select the top 10 differential expressed metabolites (DP)
DP_names_ca <- DF_stats_ca %>% 
               tibble::as.tibble() %>% 
               dplyr::select(Metabolite,'log2FoldChange', padj) %>% 
               dplyr::filter(log2FoldChange> 1.5 & 
               padj < 0.05 | log2FoldChange < -1.5 & 
               padj < 0.05) %>%
               arrange(log2FoldChange,desc(padj)) %>% 
               slice(1:10) #improved the code

DP_names_ca <- DP_names_ca$Metabolite
DF_stats_ca$dplabel <- DF_stats_ca$Metabolite
DF_stats_ca  <- DF_stats_ca %>%  as.tibble() %>% mutate(dplabel = replace(dplabel, !dplabel %in% DP_names_ca, NA) )

#generate Volcano plot for TNFa vs controls in cells (ca)

vol_ca_plot <- ggplot(data=DF_stats_ca, 
       aes(x=log2FoldChange, y=-log(padj), colour=Class, label = dplabel))+ 
    geom_point() + 
    geom_vline(xintercept=c(-1.5,1.5), color="black", alpha=0.5, linetype = "dashed")+ 
  geom_hline(yintercept=3, color="black", alpha=0.5) +
    theme_bw()+
    ggrepel::geom_text_repel() +
    ggtitle("Cells_TNFa vs Cells_Control")   


#original code from Omu but difficult to customise
#Volcano_ca <- plot_volcano(count_data = DF_stats_ca, 
                           #size = 2, 
                           #column = "Class", 
                           #strpattern = list_ca , 
                           #fill = pal, color = c(rep("black", 10)), 
                           #alpha = c(rep(0.7, 10)), shape = c(rep(21, 10))) +
                           #theme_bw() + theme(panel.grid = element_blank()) 

```
Volcano plots (for Cells Control vs Cells TGFb, NAs included)

```{r}
#generate the variables required for the volcano plot
list_cb <- unique(DF_stats_cb$Class) %>% as.vector()
pal <- viridis(option = "C", 10)

#top 10 labels
#with GG plot the amount of roles in the label you want have to still match the dataframe which you are plotting
#select the top 10 differential expressed metabolites (DP)
DP_names_cb <- DF_stats_cb %>% 
               as.tibble() %>% 
               dplyr::select(Metabolite,log2FoldChange, pval) %>% 
               dplyr::filter(log2FoldChange > 1.0 & pval < 0.05 | log2FoldChange < -1.0 & 
               pval < 0.05) %>% 
               arrange(log2FoldChange,desc(pval)) %>% 
               slice(1:10)

DP_names_cb <- DP_names_cb$Metabolite
DF_stats_cb$dplabel <- DF_stats_cb$Metabolite
DF_stats_cb  <- DF_stats_cb %>%  
                as.tibble() %>% 
                mutate(dplabel = replace(dplabel, !dplabel %in% DP_names_cb, NA) )

#generate Volcano plot for TNFa vs controls in cells (ca)

vol_cb_plot <- ggplot(data=DF_stats_cb, 
       aes(x=log2FoldChange, y=-log(padj), colour=Class, label = dplabel))+ 
    geom_point() + 
    geom_vline(xintercept=c(-1.0,1.0), color="black", alpha=0.5, linetype = "dashed")+ 
  geom_hline(yintercept=3, color="black", alpha=0.5) +
    theme_bw()+
    ggrepel::geom_text_repel() +
    ggtitle("Cells_TGFb vs Cells_Control")   


```

Generate code for heatmap (Cells : Control vs TNFa)
issues to fix with this dataset: TNFa and TGFb must be plotted seperately as the enrichments of metabolites are different from both (see the previous comparison)
```{r}
#Need to rewrite the code for the heatmap


#DF <- assign_hierarchy(count_data = DF, keep_unknowns = TRUE, identifier = "KEGG")
#DF_carbs <- subset(DF, Class == "Carbohydrates")
#listout samples which are for cells (control vs TNFa)
Cells_Control <- c("X1_001","X2_002","X3_003", "X4_004", "X5_005","X6_006", "X7_007", "X8_008")
Cells_TNFa <- c("X17_017", "X18_018","X19_019","X20_020","X21_021", "X22_022","X23_023","X24_024")


#generate first data subset of Cells samples only for TNFa and Controls
DF_sub <- DF_stats_ca %>% 
    tibble::as.tibble() %>% 
    dplyr::filter(log2FoldChange > 1.5 & 
               padj < 0.05 | log2FoldChange < -1.5 & 
               padj < 0.05) %>%
    select(Metabolite, all_of(Cells_TNFa), all_of(Cells_Control)) %>% 
    tidyr::pivot_longer(!Metabolite, names_to = "Sample", values_to = "Counts" )

DF_supp_table_1a <-  DF_stats_ca %>% 
    tibble::as.tibble() %>% 
    dplyr::filter(log2FoldChange > 1.5 &
               padj < 0.05) 
#generate a subset of the metadata to match with the sample IDs (lazy way of labelling the columns)
meta_sub <- metabolomics_metadata %>% 
    select(Sample, Treatment) 

DF_sub_ca <- inner_join(DF_sub, meta_sub, by = "Sample") %>% 
    select(Metabolite, Sample, Treatment, Counts) %>% 
    mutate(Sample = factor(Sample, levels = c("X1_001","X2_002","X3_003", "X4_004", "X5_005","X6_006", "X7_007", "X8_008", "X17_017", "X18_018","X19_019","X20_020","X21_021", "X22_022","X23_023","X24_024"))) %>% 
    mutate(Treatment = factor(Treatment, levels = c("Cells_Control", "Cells_TNFa")))
  

p1 <- ggplot(DF_sub_ca, aes(x = Sample, y = Metabolite, fill = log2(Counts))) +
  geom_tile(colour = " black") + 
    scale_fill_gradientn(colors = rev(hcl.colors(25, "BuPu")),limits = c(0,20), breaks = c(0,5,10,15,20) )+
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
    facet_grid(~Treatment, scales='free_x', space="free_x")  #use this to categorise by groups

#heatmap_carbs <- plot_heatmap(count_data = DF_sub, metadata = metabolomics_metadata, Factor = "Treatment", response_variable = "Metabolite", log_transform = TRUE, high_color = "goldenrod2", low_color = "midnightblue") + theme(axis.text.x = element_text(angle = 30, hjust=1, vjust=1, size = 6))
```

Heatmap for Cells_control vs Cels_TGFb
```{r}
#DF <- assign_hierarchy(count_data = DF, keep_unknowns = TRUE, identifier = "KEGG")
#DF_carbs <- subset(DF, Class == "Carbohydrates")
#listout samples which are for cells (control vs TNFa)
Cells_Control <- c("X1_001","X2_002","X3_003", "X4_004", "X5_005","X6_006", "X7_007", "X8_008")
Cells_TGFb <-c("X9_009","X11_011", "X12_012", "X13_013","X14_014", "X15_015", "X16_016") #remove sample 10 which is an outlier
#tried scaling but removes the signals
#DF_stats_ca[14:61] <- data.frame(scale(as.matrix(DF_stats_ca[14:61])))
#generate first data subset of Cells samples only for TNFa and Controls
DF_sub_cb <- DF_stats_cb %>% 
    tibble::as.tibble() %>% 
    dplyr::filter(log2FoldChange > 1.0 & pval < 0.05 | log2FoldChange < -1.0 & 
               pval < 0.05) %>%
    select(Metabolite, all_of(Cells_Control), all_of(Cells_TGFb)) %>% 
    tidyr::pivot_longer(!Metabolite, names_to = "Sample", values_to = "Counts" )

#generate a subset of the metadata to match with the sample IDs (lazy way of labelling the columns)
meta_sub <- metabolomics_metadata %>% 
    select(Sample, Treatment) 

DF_sub_cb <- inner_join(DF_sub_cb, meta_sub, by = "Sample") %>% 
    select(Metabolite, Sample, Treatment, Counts) %>% 
    mutate(Sample = factor(Sample, levels = c("X1_001","X2_002","X3_003", "X4_004", "X5_005","X6_006", "X7_007", "X8_008","X9_009","X11_011", "X12_012", "X13_013","X14_014", "X15_015", "X16_016"))) %>% 
    mutate(Treatment = factor(Treatment, levels = c("Cells_Control","Cells_TGFb")))
  

p2 <- ggplot(DF_sub_cb, aes(x = Sample, y = Metabolite, fill = log2(Counts))) +
  geom_tile(colour = " black") + 
    scale_fill_gradientn(colors = rev(hcl.colors(25, "BuPu")), limits = c(0,18), breaks = c(0,5,10,15,20))+
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
    facet_grid(~Treatment, scales='free_x', space="free_x")
```
 
 Heatmap with log2FC (TNFa and TGFb)
 
```{r}
DF <- assign_hierarchy(count_data = DF, keep_unknowns = TRUE, identifier = "KEGG")
#DF_carbs <- subset(DF, Class == "Carbohydrates")
#listout samples which are for cells (control vs TNFa)
#tried scaling but removes the signals
#DF_stats_ca[14:61] <- data.frame(scale(as.matrix(DF_stats_ca[14:61])))
#generate first data subset of Cells samples only for TNFa and Controls
DF_sub_ca_log <- DF_stats_ca %>% 
    tibble::as.tibble() %>% 
    dplyr::filter(log2FoldChange > 1.5 & padj < 0.05| log2FoldChange < -1.5 & 
               padj < 0.05) %>%
    select(Metabolite, log2FoldChange) 

DF_sub_cb_log <- DF_stats_cb %>% 
    tibble::as.tibble() %>% 
    dplyr::filter(log2FoldChange > 1.0 & pval < 0.05| log2FoldChange < -1.0 & 
               pval < 0.05) %>%
    select(Metabolite, log2FoldChange) 
   

#generate a subset of the metadata to match with the sample IDs (lazy way of labelling the columns)
meta_sub <- metabolomics_metadata %>% 
    select(Sample, Treatment) 

DF_sub_log <- full_join(DF_sub_ca_log, DF_sub_cb_log, by = "Metabolite") %>% 
    rename(TNFa = log2FoldChange.x, TGFb = log2FoldChange.y) %>% tidyr::pivot_longer(!Metabolite, names_to = "Sample", values_to = "log2FC" ) %>% mutate(Sample = factor(Sample, levels = c("TNFa", "TGFb")))

p3 <- ggplot(DF_sub_log, aes(x = Sample, y = Metabolite, fill = log2FC)) +
  geom_tile(colour = " black") + 
    scale_fill_gradient2(low = "#0066CC", na.value = "white", high = "#FF0000")+
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) + 
    theme_bw()

```
 

with NA metabolites filtered (maybe no need to show this to Daisy)
 
```{r}
#Need to rewrite the code for the heatmap


#DF <- assign_hierarchy(count_data = DF, keep_unknowns = TRUE, identifier = "KEGG")
#DF_carbs <- subset(DF, Class == "Carbohydrates")
#listout samples which are for cells (control vs TNFa)
Cells_Control <- c("X1_001","X2_002","X3_003", "X4_004", "X5_005","X6_006", "X7_007", "X8_008")
Cells_TNFa <- c("X17_017", "X18_018","X19_019","X20_020","X21_021", "X22_022","X23_023","X24_024")
Cells_TGFb <-c("X9_009","X11_011", "X12_012", "X13_013","X14_014", "X15_015", "X16_016")

#generate first data subset of Cells samples only for TNFa and Controls
DF_sub_2 <- DF_stats_ca %>% 
    tibble::as.tibble() %>% 
    dplyr::filter(log2FoldChange > 1.0 | log2FoldChange < -1.0 & 
               padj < 0.05) %>% filter(Class != "NA") %>%
    select(Metabolite, all_of(Cells_TNFa), all_of(Cells_Control), all_of(Cells_TGFb)) %>% 
    tidyr::pivot_longer(!Metabolite, names_to = "Sample", values_to = "Counts" )

#generate a subset of the metadata to match with the sample IDs (lazy way of labelling the columns)
meta_sub <- metabolomics_metadata %>% 
    select(Sample, Treatment) 

DF_sub_ca_2 <- inner_join(DF_sub_2, meta_sub, by = "Sample") %>% 
    select(Metabolite, Sample, Treatment, Counts) %>% 
    mutate(Sample = factor(Sample, levels = c("X1_001","X2_002","X3_003", "X4_004", "X5_005","X6_006", "X7_007", "X8_008", "X17_017", "X18_018","X19_019","X20_020","X21_021", "X22_022","X23_023","X24_024","X9_009","X11_011", "X12_012", "X13_013","X14_014", "X15_015", "X16_016"))) %>% 
    mutate(Treatment = factor(Treatment, levels = c("Cells_Control", "Cells_TNFa","Cells_TGFb")))
  

p4 <- ggplot(DF_sub_ca_2, aes(x = Sample, y = Metabolite, fill = log2(Counts))) +
  geom_tile(colour = " black") + 
    scale_fill_gradientn(colors = rev(hcl.colors(25, "BuPu")),limits = c(0,20), breaks = c(0,5,10,15,20) )  + 
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
    facet_grid(~Treatment, scales='free_x', space="free_x") 
```
Box plots

```{r}

DF_ca_box <- DF_stats_ca %>% 
    tibble::as.tibble() %>% 
    dplyr::filter(log2FoldChange > 1.0 & pval < 0.05 | log2FoldChange < -1.0 & 
               pval < 0.05) %>%
    select(Metabolite, all_of(Cells_TNFa), all_of(Cells_Control), Class) %>% 
    tidyr::pivot_longer(!c(Metabolite,Class), names_to = "Sample", values_to = "Counts" ) 

DF_ca_box_final <- inner_join(DF_ca_box, meta_sub, by = "Sample") %>% 
    select(Metabolite, Sample, Treatment, Counts, Class) %>% 
    mutate(Treatment = factor(Treatment, levels = c("Cells_Control", "Cells_TNFa"))) %>% filter(Class == 'Organic acids' |Class == 'Lipids' |Class == 'Nucleic acids') %>% group_by(Class)

DF_ca_box_final$Metabolite = factor(DF_ca_box_final$Metabolite, levels=c("arachidonic acid","adenosine" ,"adenosine-5-monophosphate" , "uridine" , "4-hydroxyphenylacetic acid", "lactic acid", "malonic acid", "parabanic acid"   ))
#Plot the box plot
ggplot(DF_ca_box_final, aes(x = Treatment, y=log(Counts), color = Treatment)) + 
  geom_boxplot() +theme_bw() + facet_wrap(Metabolite ~ Class, nrow = 2) + theme(axis.text.x = element_text(angle = 90, hjust=1))


```
 Box plot for TGFb
 
```{r}
DF_cb_box <- DF_stats_cb %>% 
    tibble::as.tibble() %>% 
    dplyr::filter(log2FoldChange > 1.0 & 
               pval < 0.05| log2FoldChange < -1.0 & 
               pval < 0.05) %>%
    select(Metabolite, all_of(Cells_TGFb), all_of(Cells_Control), Class) %>% 
    tidyr::pivot_longer(!c(Metabolite,Class), names_to = "Sample", values_to = "Counts" )

DF_cb_box_final <- inner_join(DF_cb_box, meta_sub, by = "Sample") %>% 
    select(Metabolite, Sample, Treatment, Counts, Class) %>% 
    mutate(Treatment = factor(Treatment, levels = c("Cells_Control", "Cells_TGFb"))) %>% filter(Class == 'Organic acids' | Class == 'Lipids' )

DF_cb_box_final$Metabolite = factor(DF_cb_box_final$Metabolite, levels=c("linoleic acid","palmitoleic acid" ,"4-hydroxyphenylacetic acid","malonic acid","vanillic acid", "malic acid"))

#Plot the box plot
ggplot(DF_cb_box_final, aes(x=Treatment, y=log(Counts), color = Treatment)) + 
  geom_boxplot() +theme_bw() + scale_colour_manual(values=c("#F8766D", "#7CAE00")) + facet_wrap(Metabolite ~ Class, nrow = 2) + theme(axis.text.x = element_text(angle = 45, hjust=1))
```
 

KEGG enrichment pathway analysis using FELLA 

```{r}
library(FELLA)
library(RCy3)
library(igraph)
#build a KEGG database locally using the FELLA package
graph <- buildGraphFromKEGGREST(
organism = "hsa",
filter.path = c("01100", "01200", "01210", "01212", "01230")) #this is to remove overview metabolic pathways

#store object in tmpdir on local machine
tmpdir <- paste0(tempdir(), "/my_database")
unlink(tmpdir, recursive = TRUE)

buildDataFromGraph(
keggdata.graph = graph,
databaseDir = tmpdir,
internalDir = FALSE,
matrices = "diffusion",
normality = "diffusion",
niter = 50)

#load the KEGG database as FELLA data
fella.data <- loadKEGGdata(
databaseDir = tmpdir,
internalDir = FALSE,
loadMatrix = "diffusion"
 )

#For TNFa only, only primary metabolites with kEGG terms will be included
DF_stats_ca_KEGG <- DF_stats_ca %>% 
    tibble::as_tibble() %>% 
    dplyr::filter(log2FoldChange > 1.5 | log2FoldChange < -1.5 & 
               padj < 0.05) %>% filter(KEGG != "NA") %>% dplyr::select('Metabolite', 'KEGG', 'log2FoldChange') 

compounds.TNFa <- DF_stats_ca_KEGG$KEGG


analysis.TNFa <- defineCompounds(
compounds = compounds.TNFa,
data = fella.data)

getInput(analysis.TNFa)
getExcluded(analysis.TNFa) #one compounds get excluded from the analysis 

#run diffusion
analysis.TNFa <- runDiffusion(
object = analysis.TNFa,
data = fella.data,
approx = "normality")

nlimit <- 250
vertex.label.cex <- .4
plot(
analysis.TNFa,
method = "diffusion",
data = fella.data,
nlimit = nlimit,
vertex.label.cex = vertex.label.cex)

```

Export the generated graph as igraph object to then be edited on Cytoscape
```{r}
g <- generateResultsGraph(
object = analysis.TNFa,
method = "diffusion",
nlimit = nlimit,
data = fella.data)

#generate results table
tab.all <- generateResultsTable(
method = "diffusion",
nlimit = 250,
object = analysis.TNFa,
data = fella.data)

```

for TGFb

```{r}
#For TNFa only, only primary metabolites with kEGG terms will be included
DF_stats_cb_KEGG <- DF_stats_cb %>% 
    tibble::as_tibble() %>% 
    dplyr::filter(log2FoldChange > 1.0 & pval < 0.05| log2FoldChange < -1.0 & 
               pval < 0.05) %>% filter(KEGG != "NA") %>% dplyr::select('Metabolite', 'KEGG', 'log2FoldChange')

compounds.TGFb <- DF_stats_cb_KEGG$KEGG

analysis.TGFb <- defineCompounds(
compounds = compounds.TGFb,
data = fella.data)

getInput(analysis.TGFb)
getExcluded(analysis.TGFb) #three compounds get excluded from the analysis -> include this in the background compounds.


#run diffusion
analysis.TGFb <- runDiffusion(
object = analysis.TGFb,
data = fella.data,
approx = "normality")

nlimit <- 160
vertex.label.cex <- .4
plot(
analysis.TGFb,
method = "diffusion",
data = fella.data,
nlimit = nlimit,
vertex.label.cex = vertex.label.cex)

#generate results table
tab.all.TGFb <- generateResultsTable(
method = "diffusion",
nlimit = 160,
object = analysis.TGFb,
data = fella.data)
```

